AN EFFICIENT LOCALIZED MODEL ORDER REDUCTION FRAMEWORK FOR THE SHAPE OPTIMIZATION OF ADDITIVELY MANUFACTURED LATTICE STRUCTURES\break ECCOMAS CONGRESS 2024

### 2024-08-30 paper/paper.tex
progress writing

% %%% Custom commands
\newcommand{\ie}{i.\,e.\@\xspace}
\newcommand{\eg}{e.\,g.\@\xspace}
% suffixes
\newcommand{\m}{\bm\mu}
\newcommand{\gl}{\mathrm{gl}}
\newcommand{\p}{\mathrm{p}}
\newcommand{\out}{\mathrm{out}}
\newcommand{\inrm}{\mathrm{in}}
% operators
\newcommand{\vecm}{\mathrm{vec}}
\renewcommand{\top}{\mathrm{T}}

@@

@@

\added[id=pd]{TODO: add note on notation where appropriate.}

@@

The procedure for the computation of parameter-independent deformational modes is summarized in~\cref{algo:rrfpod}.
The target tolerance $\varepsilon^{\ast}$ is used to control the quality of the approximation space (by bounding the mean $\ell^2$-error), and to compute the local tolerances $\varepsilon_{\alpha}$ that steer the range approximation of each transfer operator and the tolerance $\varepsilon_{\rho}$ for the final POD.
First, an empty snapshot set $\bm{S}$ and a training set $\bm{S}_{\mathrm{train}}$ are initialized.
The training set $\bm{S}_{\mathrm{train}}$ has to be sufficiently rich, such that it is a good representation of the parameter space, and the reader is referred to~\cite{Haasdonk2017Chapter}.
Next, for each parameter value $\m_j\in\bm{S}_{\mathrm{train}}$, the range of $\bm{T}_{\m_j}$ which is now fixed with respect to the parameter value is approximated using Algorithm 1 \textsc{AdaptiveRandomizedRangeApproximation} from~\cite{Buhr2018Randomized} and the set of snapshots is extended by Gram Schmidt orthogonalization (line 5 \& 6).
Note that the parameter value $\m$ and boundary conditions $\bm{g}$ are not varied at the same time.
Finally, the POD is computed over the snapshot set $\bm{S}$ which yields the parameter-independent space $X^n$.
The dimension of the local space $X^n$, \ie{} the number of local basis functions is denoted by $n$.

In case of inhomogeneous \textsc{Neumann} boundary conditions, after termination of~\cref{algo:rrfpod}, the \textsc{Neumann} problem~\cref{eq:neumann_problem} is solved for each $\m_j\in\bm{S}_{\mathrm{train}}$ as well.
Following compression via POD the space $X^n$ is enriched by the additional \textsc{Neumann} modes.

@@

    \State{Compute $\varepsilon_{\rho}$ and $\varepsilon_{\alpha}$ from $\varepsilon^{\ast}$ according to Theorem 10 of~\cite{Himpe2018Hierarchical}}
    \State{$\bm{S}$ $\gets\;\emptyset$}\Comment{initialize snapshot set}
    \State{$\bm{S}_{\mathrm{train}}\gets\;\{\bm\mu_1,\ldots,\bm\mu_{N_\mathrm{train}}\}$}\Comment{initialize training set}
    \For{$\bm\mu_j$ in $\bm{S}_{\mathrm{train}}$}
    \State{$\bm{R}\gets$ \textsc{AdaptiveRandomizedRangeApproximation}($\bm{T}_{\bm\mu_j}$, $\varepsilon_{\alpha}$, $n_t$, $\varepsilon_{\mathrm{algofail}}$)}
    % \State{$\bm{u}_{\mathrm{Neumann}}(\bm\mu_j)\gets \bm{T}_{\bm\mu_j}(\bm{0})$}\Comment{solve additional Neumann problem if $\varGamma_{\mathrm{N}}\ne\emptyset$}
    % \State{$\bm{R}\gets \bm{R}\cup\left(\bm{u}_{\mathrm{Neumann}}(\bm\mu_j)\right)$}
    \State{$\bm{S}\gets \mathrm{orthonormalize}\left(\bm{S}\cup\bm{R}\right)$}
    \EndFor%
    \State{$\bm{B}\gets\textsc{POD}(\bm{S}, \varepsilon_{\rho})$}
    \State{\textbf{return} {$X^n=\mathrm{span}\left(\bm{B}\right)$}}
    \EndFunction%
    \end{algorithmic} 
\end{algorithm}

% \comment[id=pd]{TODO: explain tolerances to control approximation quality. Refer to HAPOD paper and mention that $\varepsilon^{\ast}$ can be used to compute tolerances at each node of the tree (epsilon star drives the RRF and the final POD. However, this means that we have to use l2-mean error as termination criterion. Too detailed?}
% \comment[id=pd]{TODO: define notation for POD.}

@@

\subsection{Construction of a global approximation} % (fold)
\label{sub:Construction of a global approximation}
Given local approximation spaces $X^n_i$ for each target subdomain $\varOmega_i^{\p}$, the construction of a global approximation via the GFEM is discussed.
The difficulty is that the basis functions of the local spaces $X^n_i$ and $X^n_j$ of two neighboring subdomains $\varOmega_i^{\p}$ and $\varOmega_j^{\p}$ are not conforming on the shared interface $\varGamma^{\p}_{ij}=\partial\varOmega_i^{\p}\cap\partial\varOmega_j^{\p}$.
However, given a suitably\footnote{According to Definition 1 of~\cite{BABUSKA1997}.} defined partition of unity $\{{\varphi}_i\}$, the local basis functions can be included in the global FE space by multiplication with the standard FE shape functions ${\varphi}_i$, such that a conforming approximation is obtained.
The global GFEM space is defined as
\begin{equation}
	\label{eq:Xgfem}
    X_{\mathrm{GFEM}}:= \sum_{i} \varphi_iX^n_i = \left\lbrace\sum_i\varphi_i\bm\xi_i \;\big|\; \bm\xi_i\in X^n_i	\right\rbrace\,.
\end{equation}
Furthermore, the GFEM function denoted as $\bm{\psi}(\bm{x}) = \varphi(\bm{x})\bm{\xi}(\bm{x}), \bm{x}\in\omega_{\alpha}$ for a patch $\omega_{\alpha}$ is shown in~\cref{fig:gfem} to illustrate the procedure of constructing the GFEM functions.
\begin{figure}[h]
    \centering
    \subfloat[Linear hat function $\varphi(\bm{x})$]{%
          \includegraphics[height=5cm]{phi.pdf}%
          \label{fig:gfem_left}%
       } 
       \subfloat[Enrichment function $\bm{\xi}(\bm{x})$]{%
          \includegraphics[height=5cm]{enrichment.pdf}%
          \label{fig:gfem_middle}%
       }
       \subfloat[GFEM function $\bm\psi(\bm{x})$]{%
          \includegraphics[height=5cm]{gfem_function.pdf}%
          \label{fig:gfem_right}%
       }
       \caption{Construction of GFEM functions.}
       \label{fig:gfem}
\end{figure}

@@

\section{NUMERICAL EXPERIMENTS} % (fold)
\label{sec:numerical experiments}
The proposed methodology to construct the ROM is evaluated based on the example of a graded concrete slab. In~\cref{sub:Projection error study}, the approximation quality of the local spaces is accessed by studying the projection error. The ROM is validated against the FOM (\cref{sub:Reduced order model validation}) and the solution of a shape optimization problem is presented in~\cref{sub:Shape optimization}.
\comment[id=pd]{TODO: remove these sentences in case the article is too long.}

@@

\subsection{Graded concrete slab} % (fold)
\label{sub:Graded concrete slab}
The numerical example investigated here is motivated by \textit{graded concrete} in which voids are created by inserting hollow bodies into the concrete formwork with the goal of mass reduction, see~\cref{fig:gradientenbeton}.
The problem is simplified and we consider a two dimensional beam under plane stress conditions as shown in~\cref{fig:beam_sketch}.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.95\textwidth]{example-image}
    \end{center}
    \caption{}\label{fig:beam_sketch}
\end{figure}

\begin{itemize}
    \item introduce graded concrete / problem (gradientenbeton eu); why this is interesting
    \item sketch of the mechanical problem
    \item setup regarding the new method?
\end{itemize}

% \begin{figure}
%     \begin{center}
%         \includegraphics[width=0.95\textwidth]{./figures/beam/beam_sketch.pdf}
%     \end{center}
%     \caption{Schematic representation of the structural problem. A simply supported beam is subjected to a constant load.}\label{fig:beam_sketch}
% \end{figure}
%
% TODO placement
% \begin{figure}
%     \begin{center}
%         \includegraphics[width=0.65\textwidth]{./figures/beam/unit_cell.png}
%     \end{center}
%     \caption{Unit cell domain partition.}\label{fig:unit_cell_domain}
% \end{figure}
%
% \begin{figure}
%     \begin{center}
%         \includegraphics[width=0.95\textwidth]{./figures/beam/global_domain.png}
%     \end{center}
%     \caption{The partition of the global domain. The colors indicate the variation in Young's modulus for each subdomain.}\label{fig:global_domain}
% \end{figure}
%
% subsection Graded concrete slab (end)

\subsection{Projection error study} % (fold)
\label{sub:Projection error study}

% subsection Projection error study (end)

@@

@@

>>>>

% %%% Custom commands
\newcommand{\ie}{i.\,e.\@\xspace}
\newcommand{\eg}{e.\,g.\@\xspace}
% suffixes
\newcommand{\m}{\bm\mu}
\newcommand{\gl}{\mathrm{gl}}
\newcommand{\p}{\mathrm{p}}
\newcommand{\out}{\mathrm{out}}
\newcommand{\inrm}{\mathrm{in}}
\newcommand{\fom}{\mathrm{fom}}
\newcommand{\rom}{\mathrm{rom}}
\newcommand{\train}{\mathrm{train}}
% operators
\newcommand{\vecm}{\mathrm{vec}}
\renewcommand{\top}{\mathrm{T}}

@@

% \usepackage{stanli}

@@

\added[id=pd]{TODO: add note on notation where appropriate.}\\
\added[id=pd]{TODO: shorten introduction.}

@@

The procedure for the computation of parameter-independent deformational modes is summarized in~\cref{algo:rrfpod}.
The target tolerance $\varepsilon^{\ast}$ is used to control the quality of the approximation space (by bounding the mean $\ell^2$-error), and to compute the local tolerances $\varepsilon_{\alpha}$ that steer the range approximation of each transfer operator and the tolerance $\varepsilon_{\rho}$ for the final POD.
First, an empty snapshot set $\bm{S}$ and a training set $\bm{S}_{\mathrm{train}}$ are initialized.
The training set $\bm{S}_{\mathrm{train}}$ has to be sufficiently rich, such that it is a good representation of the parameter space, and the reader is referred to~\cite{Haasdonk2017Chapter}.
Next, for each parameter value $\m_j\in\bm{S}_{\mathrm{train}}$, the range of $\bm{T}_{\m_j}$ which is now fixed with respect to the parameter value is approximated using Algorithm 1 \textsc{AdaptiveRandomizedRangeApproximation} from~\cite{Buhr2018Randomized} (line 6).
If $\partial\varOmega^{\p}\cap\varSigma^{\p}_{\mathrm{N}}\ne\emptyset$, the additional \textsc{Neumann} problem~\cref{eq:neumann_problem} is solved and the solution appended to the current range $\bm{R}$ (lines 7-9).
The set of snapshots is then extended by the current range $\bm{R}$ (line 11).
Note that the parameter value $\m$ and boundary conditions $\bm{g}$ are not varied at the same time.
Finally, the POD is computed over the snapshot set $\bm{S}$ which yields the parameter-independent space $X^n$.
The dimension of the local space $X^n$, \ie{} the number of local basis functions is denoted by $n$.

@@

    \State{Compute $\varepsilon_{\rho}$ and $\varepsilon_{\alpha}$ from $\varepsilon^{\ast}$ according to Theorem 10 of~\cite{Himpe2018Hierarchical}}
    \State{$\bm{S}$ $\gets\;\emptyset$}\Comment{initialize snapshot set}
    \State{$\bm{S}_{\mathrm{train}}\gets\;\{\bm\mu_1,\ldots,\bm\mu_{N_\mathrm{train}}\}$}\Comment{initialize training set}
    \For{$\bm\mu_j$ in $\bm{S}_{\mathrm{train}}$}
    \State{$\bm{R}\gets$ \textsc{AdaptiveRandomizedRangeApproximation}($\bm{T}_{\bm\mu_j}$, $\varepsilon_{\alpha}$, $n_t$, $\varepsilon_{\mathrm{algofail}}$)}
    % \State{$\bm{u}_{\mathrm{Neumann}}(\bm\mu_j)\gets \bm{T}_{\bm\mu_j}(\bm{0})$}\Comment{solve additional Neumann problem if $\varGamma_{\mathrm{N}}\ne\emptyset$}
    % \State{$\bm{R}\gets \bm{R}\cup\left(\bm{u}_{\mathrm{Neumann}}(\bm\mu_j)\right)$}
    \If{$\partial\varOmega^{\p}\cap\varSigma^{\p}_{\mathrm{N}}\ne\emptyset$}
        \State{$\bm{u}_{\textsc{Neumann}}\gets$ Solve \cref{eq:neumann_problem} for $\m_j$ }
        \State{$\bm{R} \gets \bm{R}\cup\bm{u}_{\mathrm{Neumann}}$}
    \EndIf
    \State{$\bm{S}\gets \bm{S}\cup\bm{R}$}
    \EndFor%
    \State{$\bm{B}\gets\textsc{POD}(\bm{S}, \varepsilon_{\rho})$}
    \State{\textbf{return} {$X^n=\mathrm{span}\left(\bm{B}\right)$}}
    \EndFunction%
    \end{algorithmic} 
\end{algorithm}

@@

\subsection{Construction of a global approximation} % (fold)
\label{sub:Construction of a global approximation}
Given local approximation spaces $X^n_i$ for each target subdomain $\varOmega_i^{\p}$, the construction of a global approximation via the GFEM is discussed.
The difficulty is that the basis functions of the local spaces $X^n_i$ and $X^n_j$ of two neighboring subdomains $\varOmega_i^{\p}$ and $\varOmega_j^{\p}$ are not conforming on the shared interface $\varGamma^{\p}_{ij}=\partial\varOmega_i^{\p}\cap\partial\varOmega_j^{\p}$.
However, given a suitably\footnote{According to Definition 1 of~\cite{BABUSKA1997}.} defined partition of unity $\{{\varphi}_i\}$, the local basis functions can be included in the global FE space by multiplication with the standard FE shape functions ${\varphi}_i$, such that a conforming approximation is obtained.
The global GFEM space is defined as
\begin{equation}
	\label{eq:Xgfem}
    X_{\mathrm{GFEM}}:= \sum_{i} \varphi_iX^n_i = \left\lbrace\sum_i\varphi_i\bm\xi_i \;\big|\; \bm\xi_i\in X^n_i	\right\rbrace\,.
\end{equation}
Furthermore, the GFEM function denoted as $\bm{\psi}(\bm{x}) = \varphi(\bm{x})\bm{\xi}(\bm{x}), \bm{x}\in\omega_{\alpha}$ for a patch $\omega_{\alpha}$ is shown in~\cref{fig:gfem} to illustrate the procedure of constructing the GFEM functions.
\begin{figure}[h]
    \centering
    \subfloat[Linear hat function $\varphi(\bm{x})$]{%
          \includegraphics[height=5cm]{phi.pdf}%
          \label{fig:gfem_left}%
       }\hfill 
       \subfloat[Enrichment function $\bm{\xi}(\bm{x})$]{%
          \includegraphics[height=5cm]{enrichment.pdf}%
          \label{fig:gfem_middle}%
       }\hfill
       \subfloat[GFEM function $\bm\psi(\bm{x})$]{%
          \includegraphics[height=5cm]{gfem_function.pdf}%
          \label{fig:gfem_right}%
       }
       \caption{Construction of GFEM functions.}
       \label{fig:gfem}
\end{figure}

@@

\section{NUMERICAL EXPERIMENTS} % (fold)
\label{sec:numerical experiments}
% The proposed methodology to construct the ROM is evaluated based on the example of a graded concrete slab. In~\cref{sub:Projection error study}, the approximation quality of the local spaces is accessed by studying the projection error. The ROM is validated against the FOM (\cref{sub:Reduced order model validation}) and the solution of a shape optimization problem is presented in~\cref{sub:Shape optimization}.
% \comment[id=pd]{TODO: remove these sentences in case the article is too long.}

@@

\subsection{Graded concrete slab} % (fold)
\label{sub:Graded concrete slab}
The numerical example investigated here is motivated by \textit{graded concrete} in which voids are created by inserting hollow bodies into the concrete formwork with the goal of mass reduction, see~\cref{fig:gradientenbeton}.
The problem is simplified and we consider a two dimensional beam under plane stress conditions as shown in~\cref{fig:beam_sketch}.
The coarse grid partition of half of the beam body $\varOmega^{\p}_{\gl}=[0, 10a]\times[0, a]$ is given by $10$ quadrilateral cells of size $a$, where $a$ is the unit length of the unit cell.
Each unit cell, \ie{} subdomain $\varOmega^{\p}_i$, is taken as a square of size $a$ with a circular void with radius $R$ and is partitioned in \added{384} isoparametric quadrilateral FEs of polynomial degree two, see~\cref{fig:oversampling_domain}\,(b).
With regard to the geometrical parametrization, we consider the variation of the radius $R$ in each subdomain.
For the global problem this results in a parameter space $\mathcal{P}_{\gl}=[R_{-}, R_{+}]^{10}$, such that a parameter value $\m\in\mathbb{P}_{\gl}$ has parameter components $\m_i\in[R_{-},R_{+}], i=1, \ldots, 10$, with ranges chosen as $R_{-}=0.1a, R_{+}=0.3a$.
The point load is distributed as a constant traction over the length of one unit cell (on the far left).

\paragraph{Full order model}
An auxiliary problem is defined on the global domain $\varOmega^{\p}_{\gl}$ to account for the variation in the geometry.
The full order model is given by a standard FE model using the fine grid discretization of $\varOmega^{\p}_{\gl}$ as described above.

\paragraph{Reduced order model}
The ROM is constructed using the procedure for range approximation described in~\cref{sub:Construction of local approximation spaces}.
Using the coarse grid as above, we choose to solve $k=1, \ldots, 11$ oversampling problems, such that for each of the target subdomains $\varOmega^{\p, k}_{\inrm}$ and $\varOmega^{\p, k+1}_{\inrm}$ an overlap of size $a\times a$ is created.
The pointwise overlap condition is required for the construction of the GFEM functions.
The size of the target subdomain $\varOmega^{\p, k}_{\inrm}$ is $2a \times a$ for $k=2, \ldots, 10$ and $a \times a$ for $k = 1, 11$.
For each problem the oversampling layer has size $a \times a$, \ie{} the target subdomain is extended by a unit cell on each side unless the boundary of the target subdomain intersects with the boundary of the global domain ($\partial\varOmega^{\p,k}_{\inrm}\cap\partial\varOmega^{\p}_{\gl}\ne\emptyset$).

The input parameters to~\cref{algo:rrfpod} are $\varepsilon^{\ast}=0.001\,, n_t=\dim(\mathcal{S})$ and $\varepsilon_{\mathrm{algofail}}=1\cdot\,10^{-14}$ for $k=1, \ldots, 11$, where $\dim(\mathcal{S})$ denotes the dimension of the source space of the transfer operator and corresponds to the number of degrees of freedom on the boundary $\varGamma^{\p}_{\out}$ in the FE model.
The number of training samples is chosen as $N_{\train}=50$ ($k=1, 11$), $N_{\train}=100$ ($k=2, 10$) and $N_{\train}=200$ elsewhere, as the local parameter spaces have different dimensions.

\begin{itemize}
    \item \added{number of snapshots for each POD}
    \item difference to full order model: only use operators on unit cell domain in the online phase
\end{itemize}

@@

\begin{figure}[h]
    \centering
    \subfloat[Mechanical model of the beam]{%
          \includegraphics[keepaspectratio]{traeger.pdf}%
          \label{fig:beam_sketch_left}%
       }\hfill 
       \subfloat[Simulation model of half of the beam]{%
          \includegraphics[keepaspectratio, scale=0.8]{halfbeam.pdf}%
          \label{fig:beam_sketch_right}%
       }
       \caption{Mechanical model of the beam and representation of the simulation model with loads and kinematical constraints.}
       \label{fig:beam_sketch}
\end{figure}

% subsection Graded concrete slab (end)

@@

For the validation of the ROM, the relative error with respect to the FOM is computed over a parameter set of size $200$.
The error in the displacements in measured in the energy norm (over the domain $\varOmega^{\p}_{\gl}$) and defined as
\begin{equation}
    e_u(\m) := \frac{\norm{\bm{u}_{\fom} - \bm{u}_{\rom}}}{\norm{\bm{u}_{\fom}}}\,.
    \label{eq:error_u}
\end{equation}
The error in the first principal component of the \textsc{Cauchy} stress is defined as
\begin{equation}
    e_{\sigma}(\m) = \frac{\vert\vb{s}_{\fom} - \vb{s}_{\rom}\vert}{\vert \mathrm{max}(\vb{s}_{\fom})\vert}\,.
    \label{eq:error_sigma}
\end{equation}
Here, $\vb{s}\in\mathbb{R}^{N_g}$ denotes the vector holding the first principal stress values at the $N_g$ integration points of the mesh.
For ease of notation, the parameter-dependence is omitted for $\bm{u}$ and $\vb{s}$.
The minimum, average and maximum of $e_u$ and $e_{\sigma}$ over the validation set are shown in~\cref{fig:relerr}.

\begin{figure}
    \begin{center}
        \includegraphics[width=0.65\textwidth]{relerr.pdf}
    \end{center}
    \caption{}\label{fig:relerr}
\end{figure}

@@

\begin{itemize}
    \item problem definiot
    \item output computation (difference with ROM)
    \item result for both models
\end{itemize}
